% Research objectives related to the obtained results and whether the main goal of the project was achieved
% Future work perspectives and new research opportunities
% Potential limitations that constrained the work

\chapter{Conclusion and Future Work}

\ac{LLM}s are increasingly used in agentic settings, where they act on a user's behalf, interact with other agents and make decisions under uncertainty. Negotiation is a particularly attractive scenario for studying these capabilities. A good negotiator must interpret the intentions of the other party, plan several moves ahead, and act on incomplete information, yet, unlike many agentic tasks, performance can be measured in a concrete way. NegotiationArena \cite{bianchi_how_2024} offered a framework to do exactly that, but its original evaluation covered only proprietary models. This raised concerns about reproducibility and accessibility, and left open the question of how smaller open-weight models behave in the same scenarios.

This thesis set out to answer that question and to test whether Self-Refine and team negotiation can make these models negotiate better. We started by cross-playing three model families (Gemma, Mistral and Qwen) across three tiers, comparing their behaviours against those reported for proprietary models. Then we replicated the social-persona experiment from NegotiationArena, assigning one party a \textit{desperate} or \textit{cunning} persona, before measuring the effect of Self-Refine and team negotiation. The remainder of this chapter summarises what we learned, reflects on the difficulties encountered along the way (Section~\ref{sec:conclusion-reflections}), states the limitations that constrained the study (Section~\ref{sec:conclusion-limitations}) and outlines directions for future work (Section~\ref{sec:conclusion-future-work}).

\paragraph{Open-weight models are capable negotiators.} The nine models we tested are able to play all three games, and they reproduce several of the patterns that \textcite{bianchi_how_2024} reported for proprietary models. We observe the same price anchoring in BuySell (a 0.752 correlation between the seller's opening and the final agreed price, close to the 0.716 they report) and the same positional asymmetries: the second mover wins, on average, 47.7 percentage points more often in Trading and 53.9 in BuySell, while the proposer wins 81.5 points more often in the Ultimatum Game. Like proprietary models, they also behave irrationally at times, opening below the production cost in 16.2\% of BuySell games and rejecting Multi-turn Ultimatum offers outright instead of countering (a value-destroying move for both). Across families, Qwen is the most consistent negotiator, winning 63--67\% of games, ahead of Gemma and Mistral, though in the Ultimatum Game Gemma is the strongest family.

\paragraph{Social personas do not transfer cleanly from proprietary models.} \textcite{bianchi_how_2024} found that giving Player 2 a \textit{cunning} or \textit{desperate} persona raised both its win rate and payoff across all three games in GPT-4 self-play. This does not generalize to open-weight models: the effect is conditional on the game and on the model carrying it. \textit{Cunning} lifts the Ultimatum responder's win rate from 14.3\% to 70.5\%, but it also increases the no-deal rate to 73\% (from 13\% under default) and returns the lowest mean payoff of the three conditions. \textit{Desperate} raises BuySell buyer's win rate from 75.2\% to 96.7\% and talks sellers below their production cost. Neither persona helps in Trading. The size of the effect also varies sharply from one model to the next, for example: being cunning in the Ultimatum Game shifts win rate by +0.90 for \texttt{Gemma-3-27B-it} but by -0.47 for \texttt{Ministral-3-8B}. Personas amplify a style whose value depends on the game and on the model carrying and facing it, rather than reliably improving negotiation.

\paragraph{Self-Refine is not a general improvement.} Adding a draft-critique-rewrite loop does not make open-weight models negotiate better in general. Its effect is \textit{asymmetric}: it tends to help a party that starts from a structurally weak position, improving the disadvantaged seat in 7 of 9 configurations (a mean +5.7 percentage points of win rate, most clearly the Trading first mover at +14.5 to +19.0 points), while penalizing seats that are already strong, such as the medium-tier BuySell seller, which loses 20.3 points. Only 5 of the 18 configurations are statistically significant, and where the loop does change a move, it hardens it (claims more) more often than it softens it. None of this justifies the cost: a refined move generates roughly six times as much text as a default one (a median of 4,900 characters against 790). A plausible reason for the lack of performance is that, unlike the coding and math tasks Self-Refine was designed for, a negotiation has no verifiable target for the critique to converge on.

\paragraph{Team negotiation does not justify its cost.} Replacing one party with a deliberating team of three models, evaluated at the 12--14B tier, leaves win rate essentially unchanged: not a single team-versus-single-agent comparison has separated confidence intervals, and the average lift is only 0.035. Payoff gains are localized, appearing on structurally weak seats (the heterogeneous BuySell seller gains 5.6ZUP). We found that heterogeneous teams produce more diverse drafts than homogeneous ones ($\approx 0.50$ versus $\approx 0.36$ mean pairwise cosine distance), but this diversity does not translate into better outcomes. One likely reason is that across heterogeneous teams, the Ministral draft wins the Borda count 80.5\% of the time, and all three families rank it first most often, even though Qwen was the strongest single model in the benchmark. For an order-of-magnitude increase in cost (12--15x the number of characters), team deliberation is not a reliable improvement at this tier.


\section{Reflections}\label{sec:conclusion-reflections}

Our first practical obstacle was the fragility of the smaller open-weight models with respect to the response protocol. A single missing or malformed \ac{XML} tag is enough to end a game prematurely, and at the very-small tier this happened often enough to threaten the validity of the comparison. This motivated the parse-error self-correction loop (Section~\ref{sec:retry-ablation}), which recovers most malformed moves. Building and validating that loop was an important part of this thesis.

A second difficulty was the sheer size of the experimental space and the variance it produced. With three families, three tiers, three games, two seats and several conditions per technique, analysing results required care. Pooling results, reporting confidence intervals and repeatedly cross-checking our observations against those reported by \textcite{bianchi_how_2024} were essential to separate effects from noise.

Running the experiments was itself a challenge. We had to run inference across three different environments, two SLURM clusters and Kaggle's GPUs, each with its own quotas, hardware, and quirks. Orchestrating the sweeps and reconciling the resulting logs into a single dataset took a substantial share of the effort behind this thesis.

\section{Limitations}\label{sec:conclusion-limitations}

The main limitation of this work was the computational resources available, which shaped many of the decisions reported in the previous chapters. Most of our experiments were run on the shared MIA cluster. Because it is shared across a considerable number of students, it was frequently busy, and we could not always launch a job or iterate on an idea immediately. This led us to look for alternatives. We requested access to the Deucalion supercomputer, which helped considerably, but our allocation was limited to a budget of 150 GPU hours. We later discovered that Kaggle's command-line interface was a valuable third option: although its free GPUs could only host the smaller models, they let us run small iterations almost immediately, within the limits of two concurrent jobs and thirty GPU hours per week. Together these three environments made this study feasible, but the constant struggle with queues, quotas, and budgets bounded how much we could ultimately run.

That constraint is most visible in the number of runs. Because the thirty games per cell were spread across three families, three tiers, two seats and several conditions per technique, many of our confidence intervals are wide. We were also bounded on the models we could run. We covered three open-weight families at tiers up to 24--27B, but did not evaluate larger or frontier open-weight models such as DeepSeek or Kimi. Quantization made the larger tiers tractable and quicker to run but may not perfectly reflect the behaviour of the full-precision weights. The team-negotiation experiments were run only at the 12--14B tier, leaving open whether the consensus bias towards the Ministral draft would persist at other model sizes.

Two further limitations are methodological rather than computational. To keep the comparison with NegotiationArena fair, we took the game prompts unchanged from \textcite{bianchi_how_2024} and did not tune them per model family; a model that parses or negotiates poorly under these prompts might do better under prompts adapted to it. Finally, our study of Self-Refine was fixed to have a single critique and two refinement iterations, so the mixed result we report speaks only to this particular configuration.

\section{Future Work}\label{sec:conclusion-future-work}

The findings above, and the limitations just described, suggest several directions that we believe would strengthen this line of work:

\begin{itemize}
    \item \textbf{More realistic negotiation scenarios:} the NegotiationArena games are deliberately simple. Richer settings, with more issues to trade off and concrete goods and currencies in place of the abstract resources (X, Y, ZUP), would allow us to test LLMs on conditions closer to real negotiation.
    \item \textbf{Increase the number of runs:} re-running the benchmark with more games per cell would narrow the confidence intervals and allow for more robust statistical conclusions.
    \item \textbf{Cross-play with proprietary and frontier open-weight models:} matching open-weight models directly against proprietary ones, and adding stronger open-weight models such as DeepSeek or Kimi, would better quantify the gap between open-weight and proprietary models.
    \item \textbf{Varying game parameters:} all games used the fixed setups inherited from NegotiationArena (seller cost, buyer value, pot size) for direct comparison. Varying these parameters would be a natural next step.
    \item \textbf{Negotiation styles:} beyond the two NegotiationArena personas, future work could study the impact of specific negotiation tactics such as good-cop/bad-cop.
\end{itemize}

\textbf{Self-Refine:}
\begin{itemize}
    \item \textbf{Critique design:} our critique used a fixed set of criteria. Game-specific or alternative critique prompts might help in cases where the generic critique did not.
    \item \textbf{Adaptive stopping:} rather than always running two iterations, the loop could stop once a draft stops changing, or once a cheap heuristic judges it good enough, reducing some of the technique's cost.
\end{itemize}

\textbf{Team Negotiation:}
\begin{itemize}
    \item \textbf{Explaining the Ministral preference:} as discussed in Section~\ref{sec:consensus-selection}, in heterogeneous teams, all three families ranked the Ministral draft first far more often than any other, even though Qwen was the stronger model. We ruled out candidate order and draft length, but the underlying cause remains open. Identifying what drives this bias, and whether it persists across model tiers, would clarify whether team deliberation suffers from a general ``expertise dilution'' effect.
    \item \textbf{Alternative consensus mechanisms:} the Borda count is one of several consensus mechanisms studied by \textcite{S_nchez_Anguix_2013}. Comparing it against majority voting, confidence-weighted voting in the style of RECONCILE \parencite{chen2024reconcileroundtableconferenceimproves}, a single representative deciding for the team, or a neutral judge could improve the negotiation team's performance.
    \item \textbf{Explicit member roles and topology:} we deliberately kept the team symmetric and fully connected. Assigning roles such as leader, analyst, or devil's advocate, and structuring the discussion as a star, chain, or mesh, may increase the diversity that our current implementation lacks.
    \item \textbf{Team size:} we fixed the team at three members; varying it would let us study the trade-off between diversity of proposals and computational cost.
\end{itemize}